# Network data: how to store and use network data {#nwdata}
## Different formats to store network data

```{r include=FALSE}
knitr::opts_chunk$set(fig.path = 'mainimages/')
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
library(statnet)
library(ggplot2)
library(GGally)
library(scales)
```

Relational data consist of nodes and edges between these nodes. Data on the nodes---nodal attributes---can be stored in the conventional way of storing observational data: in a data frame where each row corresponds to each node (observation) in the network and columns to variables (eg. gender). 
The edges that make up a network need to be stored in a slightly more complex manner. 

Take the following network of friendship relations between members of the Swiss national soccer team, leading up to the 2014 world championship. Players were asked in a series of interviews with each player in Neue Zürcher Zeitung about their friends in the team. This information was used to create a network dataset for illustrative purposes. https://www.nzz.ch/sport/wm-2014/portraetserie-der-schweizer-nationalspieler-1.18315579

```{r, echo=FALSE}
myel <- read.csv("bookdown_resources/nati_friends.csv", stringsAsFactors = FALSE, encoding = "UTF-8")
myel <- myel[myel$Receiver != "Alle",]

## create adjacency matrix

unique_names <- unique(c(myel$Sender, myel$Receiver))
friends_adjmat <- matrix(0, nrow = length(unique_names), ncol = length(unique_names))
colnames(friends_adjmat) <- unique_names
rownames(friends_adjmat) <- unique_names
friends_adjmat[cbind(myel$Sender,myel$Receiver)] <- 1

att <- read.csv("bookdown_resources/nati_atts.csv", stringsAsFactors = FALSE, encoding = "UTF-8")
att <- att[att$Name %in% unique_names,]
att <- att[match(att$Name, unique_names),]
```

```{r trialfriendsadjmat, echo=FALSE, fig.cap='Friendship relations between members of the Swiss national soccer team'}
## plot the network
ggnet2(friends_adjmat,
       label = TRUE,
       arrow.size = 4, arrow.gap = 0.04, 
       node.color = att$Position) 
```

To store the network data leading to the network plot, the sender and target nodes of each edge in the network have to be recorded. There are multiple ways of storing network data. The simplest form is called a `edgelist`, where each row in a data set corresponds to a tie in the network: 

```{r, echo=FALSE}
myel
```

Here the rows correspond to the edges in the network, the columns correspond to sender nodes `myel[,1]` and target nodes `myel[,2]`.

An alternative way of storing network data is by creating an adjacency matrix with all unique nodes in the rows as well as in the columns: 

```{r, echo=FALSE}
friends_adjmat[1:10,1:10]
```

Here, every possible tie in a network is listed in the data set, for instance sender `a` tying to target `b`. If this tie exists, if `(a, b)` are indeed linked in the network, the matrix alotts them a 1. If `(a, b)` are not linked, they receive a 0 in the adjacency matrix. This way of storing network data also preserves the direction of the tie. Entries in the lower triangle of the matrix specify a tie from row entry to column entry and vice versa for the upper triangle. For large networks, storing data in adjacency matrices can become inefficient, as a lot of information is redundant (non-existing ties). However, many ways of computing network statistics operate directly on the adjaceny matrix.

A third way of storing network data is through adjacency lists. Adjacency list specify a sender first and then list all receivers. They are the most efficient format for gathering data in the field or transcribing/ entering data by hand.

```{r, echo = FALSE}
adj_list <- apply(rbind(colnames(friends_adjmat),friends_adjmat), 2, function(x) ifelse(x == 1,x[1],""))
adj_list <- cbind(rownames(adj_list),apply(adj_list, 1, function(x) paste(x[x !=""], collapse = ",")))
rownames(adj_list) <- NULL
adj_list <- adj_list[-1,]
```


As mentioned above, nodal attributes can be stored in a data frame with row corresponding to unique nodes and columns to variables: 

```{r, echo=FALSE}
att
```

## Dealing with network data in R

### Reading in data stored in an edge list

Since edge lists (or adjacency lists) are the most efficient way of storing your data, we urge you to compile your network data in excel/txt-file using the sender-target columns for individual network ties.

Reading in your data: 
```{r, eval=FALSE}
myel <- read.table(file = "myedgelist.csv", sep = ";", header = TRUE)
```


```{block, type ='rmdtask'}
Read in the network data of the Swiss national socer team.
```


### Converting an edge list to adjacency matrix

Next you'll need to determine how many unique nodes you have in your network:  

```{r, eval = FALSE}
mynodes <- unique(c(myel$sender, myel$target))
```

Once you have specified a vector with all the unique names of the nodes in your network, you can create an empty matrix (empty = filled with zeros) and label the columns and rows accordingly: 

```{r, eval=FALSE}
mat <- matrix(0, nrow = length(mynodes), ncol = length(mynodes))
colnames(mat) <- mynodes
rownames(mat) <- mynodes
```

Next you can fill your empty matrix with the ties reported in your egelist: 

```{r, eval=FALSE}
mat[cbind(myel$sender, myel$target)] <- 1
```

### Converting an adjacency list to an edge list

If you have collected your data as an adjacency list, you can convert it to an edge list using the procedure below. It requires that your adjacency list is stored as a matrix with the first column specifing the sender and the remaining columns specifying receivers:

```{r, echo=FALSE}
vec_list <- unlist(lapply(adj_list[,2], strsplit, split = ","),recursive = F)
vec_list[unlist(lapply(vec_list, function (x) length(x))) == 0] <- NA
max_adjlist <- max(unlist(lapply(vec_list, length)))
adjlist_mat <- do.call(what = "rbind",lapply(vec_list, function(x) {na_vec <- rep(NA,max_adjlist)
na_vec[1:length(x)] <- x
na_vec}))
adjlist_mat <- cbind(adj_list[,1], adjlist_mat)
```

```{r}
adjlist_mat
```

The procedure is slightly complicated-looking, but it's alright :).

```{r}
el <-
  cbind(
    c(apply(adjlist_mat,1,function(x) rep(x[1], length(x)-1))),
    c(apply(adjlist_mat,1,function(x) x[2:length(x)]))  
  )
el <- el[!(is.na(el[,2])),]
el
```


### Storing nodal attributes the right way

Once you have prepared your adjacency matrix, you can create your nodal attribute data frame. 

```{block, type ='rmdwrong'}
Never use an attribute data frame without having made sure that the rownames of your adjacency matrix match the rows in your attribute file!
```

You can either create a new data frame that contains a variable with the unique nodes, or you can sort a given attribute file so that the rows reflect the same people in the attribute file as well as in the adjacency matrix.

```{r, eval=FALSE}
dt <- read.table(file = "attributes.csv", sep = ";", header = TRUE)
```

Let's assume your data frame contains a variable `nodeID` with all unique node names. Then you can check if your attribute file matches the adjacency matrix:

```{r, eval = FALSE}
identical(att$nodeID, rownames(mat))
```

If they match, you can proceed. If they don't match, you'll have to sort your data: 

```{r, eval=FALSE}
att <- att[att$nodeID %in% mynodes,]
## or: 
att <- att[att$nodeID %in% rownames(mat),]
## or: 
att <- att[match(att$nodeID, mynodes),]
## or: 
att <- att[match(att$nodeID, rownames(mat)),]
```

or create a new data frame: 

```{r, eval=FALSE}
att <- data.frame(nodeID = rownames(mat))
```

and then match variables from your original attributes file to your new attribute file: 

```{r, eval=FALSE}
att$newVariable <- dt$variable[match(att$nodeID, dt$nodeID)]
```


### The network object from the `statnet`-package

The `statnet`-package mainly works with a `network`-object. This means that once you have prepared your adjacency matrix (here named `mat`) and your attributes file (here named `att`) you have to create a nework object with which you can then work with. 

```{r, eval=FALSE}
nw <- network(mat, directed = TRUE)
```

The `network()`-command takes various options. If you have a directed network, specify `directed = TRUE`, otherwise `directed = FALSE`.

The network object stores all your information on the network: the number of nodes, the number of edges, whether it's a one-mode or two-mode network, whether it's directed or not, etc. 

For the Swiss national soccer team friendship network, the object looks like this: 

```{r, echo=FALSE}
network(friends_adjmat, directed = TRUE)
```

You can see that the network contains 19 nodes (or vertices) and 27 edges (i.e., friendship ties).

It can also store your network attributes that you have saved in the `att`-data frame until now: 

```{r, eval = FALSE}
set.vertex.attribute(nw, 'gender', att$gender)
## or: 
nw %v% 'gender' = att$gender
```

Now the variable `att$gender` is stored in the network `nw` and is labeled `gender`. You can give it whichever label you'd like, but make sure there are no spaces in between (this may cause errors later on).

## Different Network types: one-mode and two-mode networks

Two-mode networks consist of two disjoint sets of nodes (modes) with relations only possible between modes. For example, a two-mode network could illustrate associations between Swiss national soccer players and the clubs they played in between 2012 and 2014.

```{r, echo=FALSE}
twomode_el <- read.csv("bookdown_resources/nati_teams.csv", stringsAsFactors = FALSE, encoding = "UTF-8")

unique_players <- unique(twomode_el$Player)
unique_teams <- unique(twomode_el$Team)
bip_mat <- matrix(0, nrow = length(unique_players), ncol = length(unique_teams))
colnames(bip_mat) <- unique_teams
rownames(bip_mat) <- unique_players
bip_mat[cbind(twomode_el$Player,twomode_el$Team)] <- 1

col = c("actor" = "forest green", "event" = "light blue")
ggnet2(network(bip_mat, matrix.type = "bipartite"), 
       label = TRUE, palette = col,
       color = "mode", label.size = 2.5)
```

A matrix storing ties for a two-mode network is rectangular and commonly referred to as an incidence or occurence matrix.

```{r, echo=FALSE}
bip_mat[1:4,1:7]
```

A projection onto a one-mode network can be achieved by multiplying the rectangular matrix by its transpose. Now players are connected based on whether they played together in the same clubs.

```{r}
one_mode_projection <- bip_mat %*% t(bip_mat)
diag(one_mode_projection) <- 0 #exclude self-ties
one_mode_projection[1:10,1:10] #look at the first 10 players
```

```{r, echo=FALSE}
ggnet2(network(one_mode_projection),
       label = TRUE)
```

## Network visualization

### A word of warning

Network visualizations are at the same time pretty and dangerous. Be aware that nodes will always be placed on a canvas through an algorithm. The choice of a different algorithm can lead to a vastly different illustration. It is thus easy to convey substantive results that are just artefacts of the visualization procedure, for example regarding the centrality of actors. It is in your responsibility to use visualization to augment, not distort interpretation. If you read a network study engaging in "visual interpretation" - look for the numbers. Do they back up the story?

### Resources for visualization

For pretty graphs in R, ggnet2 is hard to beat. https://briatte.github.io/ggnet/

For more advanced (and even prettier) visualizations, Gephi is a dedicated open source project (although somewhat buggy). https://gephi.org/

A third, java-based visualization tool is Visone. It is hard to beat when you want to look at different clusters in your network (but contains almost all other network visualization tools as well). https://visone.info

### A brief overview over ggnet2()

The command `ggnet2()` can work with either the adjacency matrix and the attributes-data frame or with the network object. 

```{r}
ggnet2(friends_adjmat,
       label = TRUE,                     # should nodes be labeled?
       arrow.size = 4, arrow.gap = 0.04, # set arrowheads
       node.color = att$Position,        # specify node colors
       palette = 'Spectral')             # use a pretty color-palette
```

Since all `ggnet()`-plots are `ggplots` you can add theme options that you usually use in ggplots: 

```{r}
ggnet2(friends_adjmat,
       label = TRUE,                     
       arrow.size = 4, arrow.gap = 0.04, 
       node.color = att$Position,        
       palette = 'Spectral') +   
  theme(legend.position = 'bottom') +
  ggtitle("Friendship network of the Swiss national soccer team")
```

Plus you can save the plots using `ggsave()`:

```{r, eval=FALSE}
ggsave(file = 'figures/nw_soccerCH.pdf', width = 15, height = 12, units = 'cm')
```
