# Network data: how to store and use network data {#nwdata}
## Different formats to store network data

```{r include=FALSE}
knitr::opts_chunk$set(fig.path = 'mainimages/')
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
library(statnet)
library(ggplot2)
library(GGally)
library(scales)
```

Relational data consist of nodes and edges between these nodes. Data on the nodes---nodal attributes---can be stored in the conventional way of storing observational data: in a data frame where each row corresponds to each node (observation) in the network and columns to variables (eg. gender). 
The edges that make up a network need to be stored in a slightly more complex manner. 

Take the following network of friendship relations between members of the Swiss national soccer team, leading up to the 2014 world championship. Players were asked in a series of interviews with each player in Neue Zürcher Zeitung about their friends in the team. This information was used to create a network dataset for illustrative purposes. https://www.nzz.ch/sport/wm-2014/portraetserie-der-schweizer-nationalspieler-1.18315579

```{r, echo=FALSE}
myel <- read.csv("bookdown_resources/nati_friends.csv", stringsAsFactors = FALSE, encoding = "UTF-8")
myel <- myel[myel$Receiver != "Alle",]

## create adjacency matrix

unique_names <- unique(c(myel$Sender, myel$Receiver))
friends_adjmat <- matrix(0, nrow = length(unique_names), ncol = length(unique_names))
colnames(friends_adjmat) <- unique_names
rownames(friends_adjmat) <- unique_names
friends_adjmat[cbind(myel$Sender,myel$Receiver)] <- 1

att <- read.csv("bookdown_resources/nati_atts.csv", stringsAsFactors = FALSE, encoding = "UTF-8")
att <- att[att$Name %in% unique_names,]
att <- att[match(att$Name, unique_names),]
```

```{r trialfriendsadjmat, echo=FALSE, fig.cap='Friendship relations between members of the Swiss national soccer team'}
## plot the network
ggnet2(friends_adjmat,
       label = TRUE,
       arrow.size = 4, arrow.gap = 0.04, 
       node.color = att$Position) 
```

To store the network data leading to the network plot, the sender and target nodes of each edge in the network have to be recorded. There are multiple ways of storing network data. The simplest form is called a `edgelist`, where each row in a data set corresponds to a tie in the network: 

```{r, echo=FALSE}
myel
```

Here the rows correspond to the edges in the network, the columns correspond to sender nodes `myel[,1]` and target nodes `myel[,2]`.

An alternative way of storing network data is by creating an adjacency matrix with all unique nodes in the rows as well as in the columns: 

```{r, echo=FALSE}
friends_adjmat[1:10,1:10]
```

Here, every possible tie in a network is listed in the data set, for instance sender `a` tying to target `b`. If this tie exists, if `(a, b)` are indeed linked in the network, the matrix alotts them a 1. If `(a, b)` are not linked, they receive a 0 in the adjacency matrix. This way of storing network data also preserves the direction of the tie. Entries in the lower triangle of the matrix specify a tie from row entry to column entry and vice versa for the upper triangle. For large networks, storing data in adjacency matrices can become inefficient, as a lot of information is redundant (non-existing ties). However, many ways of computing network statistics operate directly on the adjaceny matrix.

A third way of storing network data is through adjacency lists. Adjacency list specify a sender first and then list all receivers. They are the most efficient format for gathering data in the field or transcribing/ entering data by hand.

```{r, echo = FALSE}
adj_list <- apply(rbind(colnames(friends_adjmat),friends_adjmat), 2, function(x) ifelse(x == 1,x[1],""))
adj_list <- cbind(rownames(adj_list),apply(adj_list, 1, function(x) paste(x[x !=""], collapse = ",")))
rownames(adj_list) <- NULL
adj_list[-1,]
```


As mentioned above, nodal attributes can be stored in a data frame with row corresponding to unique nodes and columns to variables: 

```{r, echo=FALSE}
att
```

## Dealing with network data in R

### Storing network data using edge lists





```{block, type ='rmdtask'}
Read in the network data of the Swiss national socer team.
```

### Converting an edge list to adjacency matrix



### Converting an adjacency list to adjacency


### Storing nodal attributes the right way



### The network object from the `statnet`-package



## Different Network types: one-mode and two-mode networks

Two-mode networks consist of two disjoint sets of nodes (modes) with relations only possible between modes. For example, a two-mode network could illustrate associations between Swiss national soccer players and the clubs they played in between 2012 and 2014.

```{r, echo=FALSE}
twomode_el <- read.csv("bookdown_resources/nati_teams.csv", stringsAsFactors = FALSE, encoding = "UTF-8")

unique_players <- unique(twomode_el$Player)
unique_teams <- unique(twomode_el$Team)
bip_mat <- matrix(0, nrow = length(unique_players), ncol = length(unique_teams))
colnames(bip_mat) <- unique_teams
rownames(bip_mat) <- unique_players
bip_mat[cbind(twomode_el$Player,twomode_el$Team)] <- 1

col = c("actor" = "forest green", "event" = "light blue")
ggnet2(network(bip_mat, matrix.type = "bipartite"), 
       label = TRUE, palette = col,
       color = "mode", label.size = 2.5)
```

A matrix storing ties for a two-mode network is rectangular and commonly referred to as an incidence or occurence matrix.

```{r, echo=FALSE}
bip_mat[1:4,1:7]
```

A projection onto a one-mode network can be achieved by multiplying the rectangular matrix by its transpose. Now players are connected based on whether they played together in the same clubs.

```{r}
one_mode_projection <- bip_mat %*% t(bip_mat)
diag(one_mode_projection) <- 0 #exclude self-ties
one_mode_projection[1:10,1:10] #look at the first 10 players
```

```{r}
ggnet2(network(one_mode_projection),
       label = TRUE)
```


## Network visualization

### A word of warning

Network visualizations are at the same time pretty and dangerous. Be aware that nodes will always be placed on a canvas through an algorithm. The choice of a different algorithm can lead to a vastly different illustration. It is thus easy to convey substantive results that are just artefacts of the visualization procedure, for example regarding the centrality of actors. It is in your responsibility to use visualization to augment, not distort interpretation. If you read a network study engaging in "visual interpretation" - look for the numbers. Do they back up the story?

### Resources for visualization

For pretty graphs in R, ggnet2 is hard to beat. https://briatte.github.io/ggnet/

For more advanced (and even prettier) visualizations, Gephi is a dedicated open source project (although somewhat buggy). https://gephi.org/


